---
title: " Categorical Data – MSDA IFP"
subtitle: "Week 09 – Tutorial 02"
format:
  html:
    layout: article
    toc: true
    toc-depth: 3
    code-annotations: hover
    code-line-numbers: false
    code-fold: false
    code-fold-label: "Code"
    code-block-background: true
    code-block-bg: true
    highlight-style: "pygments"
    css: styles.css
editor: 
  markdown: 
    wrap: 72
---

## 1. Tasks For formative report

This week’s task is highlighted in bold below. Please only focus on
completing that task this week. In the next section, you will also find
guided sub-steps you may want to consider to complete this week’s task.

::: task-block
<p class="task-main">

1) Read the College dataset into R, inspect it, and write a concise
introduction to the data and its structure.

</p>

<p class="task-sub">

<strong style="color:#1a73e8;">2) Display and describe the categorical variables.</strong>

</p>

<p class="task-sub">

3) Display and describe a selection of numeric variables.

</p>

<p class="task-sub">

4) Display and describe at least one relationship between two or three variables.

</p>

<p class="task-sub">

5) Finish the report write-up, knit to PDF, and submit.

</p>
:::


This tutorial is designed to help you complete **Task 2**.

## 1.1 Task 2 – sub-tasks

## 1.1 Task 2 – sub-tasks

::: {.callout-tip .margin}
**Tip:** Hover over the footnotes for hints showing useful R functions.
:::

This week you will only focus on **Task 2: Display and describe the categorical variables**.  
Below there are some guided sub-steps you may want to consider to complete **Task 2**.

- Reopen last week’s `.qmd` file, as you will continue last week’s work and build on it.[^1]

1. Read the **CollegeScores** dataset into R (from the provided CSV file in `datasets/`) and give the object a sensible name (for example `college`).[^2]

2. Use functions such as `glimpse()` or `str()` to check the structure of the dataset and identify which variables are **categorical** (e.g. `State`, `Control`, `Region`, `Locale`, `Main`, `Online`).[^3]

3. Convert these categorical variables to **factors** in R so that they are treated correctly in plots and summaries.[^4]

4. Create **bar plots** showing the frequency distribution of at least three categorical variables  
   (for example: number of institutions by `Region`, by `Control`, and by `Locale`).[^5]

5. Improve the readability of your bar plots by:
   - rotating x-axis labels if the names are long  
   - ordering categories by frequency using `fct_infreq()` where appropriate  
   - adding clear axis labels and an informative title with `labs()`.[^6]

6. For the same categorical variables, create **frequency tables** showing both the **count** and the **percentage** of institutions in each category.[^7]

7. Format your frequency tables neatly using `kbl()` from the **kableExtra** package, so they are ready to include in your report.[^8]

8. In your written **Analysis** section, describe the distributions in words. For example:
   - Which categories are most common?  
   - Are some categories rare?  
   - Is the distribution fairly balanced or dominated by one or two groups?

9. For each categorical variable in your report, include **either** the bar plot **or** the frequency table (not both) to avoid duplicating information.

[^1]: Hint: Open the same `.qmd` file you used for Task 1 so you keep building a single report.
[^2]: Hint: Use `read_csv("datasets/CollegeScores_teaching.csv")` from the **readr** package.
[^3]: Hint: Try `glimpse(college)` from **dplyr** or `str(college)`.
[^4]: Hint: Use `mutate()` with `factor()`, for example  
      `college <- college |> mutate(Control = factor(Control), Region = factor(Region))`.
[^5]: Hint: Use `ggplot(college, aes(x = Control)) + geom_bar()`.
[^6]: Hint: `theme(axis.text.x = element_text(angle = 45, hjust = 1))` and `fct_infreq()` from **forcats**.
[^7]: Hint: `college |> count(Control) |> mutate(Percent = round(n / sum(n) * 100, 2))`.
[^8]: Hint: `kbl(your_table, booktabs = TRUE, digits = 2)`.



You do **not** need to include R output in the report; you only need the
**written description**. Keep this paragraph; you will re-use it in your
report.


## Data

[CollegeScores Dataset]{.data-title}

You will work with the *CollegeScores* dataset in this tutorial.\
At the link **CollegeScores_teaching.csv** you will find information
about **400 higher-education institutions in the United States**.\
The dataset includes variables describing each institution’s location,
sector (public/private), tuition costs, enrolment, and the demographic
composition of their student body.

```{r college-direct, echo=FALSE, message=FALSE, warning=FALSE}

library(tibble)
library(gt)
library(dplyr)

college_dict <- tribble(
  ~Variable,    ~Description,
  "Name",       "Name of the school",
  "State",      "State where the school is located",
  "ID",         "ID number for school",
  "Main",       "Main campus? (1 = yes, 0 = branch campus)",
  "Accred",     "Accreditation agency",
  "MainDegree", "Predominant undergrad degree (0 = not classified, 1 = certificate, 2 = associate, 3 = bachelors, 4 = only graduate)",
  "HighDegree", "Highest degree (0 = no degrees, 1 = certificate, 2 = associate, 3 = bachelors, 4 = graduate)",
  "Control",    "Control of school (Private, Profit, Public)",
  "Region",     "Region of country (Midwest, Northeast, Southeast, Territory, West)",
  "Locale",     "Locale (City, Rural, Suburb, Town)",
  "Latitude",   "Latitude",
  "Longitude",  "Longitude",
  "AdmitRate",  "Admission rate",
  "MidACT",     "Median of ACT scores",
  "AvgSAT",     "Average combined SAT scores",
  "Online",     "Only online (distance) programmes",
  "Enrollment", "Undergraduate enrolment",
  "White",      "Percent of undergraduates who report being white",
  "Black",      "Percent of undergraduates who report being black",
  "Hispanic",   "Percent of undergraduates who report being Hispanic",
  "Asian",      "Percent of undergraduates who report being Asian",
  "Other",      "Percent of undergraduates who do not report one of the above",
  "PartTime",   "Percent of undergraduates who are part-time students",
  "NetPrice",   "Average net price (cost minus aid)",
  "Cost",       "Average total cost for tuition, room, board, etc.",
  "TuitionIn",  "In-state tuition and fees",
  "TuitionOut", "Out-of-state tuition and fees",
  "TuitionFTE", "Net tuition revenue per FTE student",
  "InstructFTE","Instructional spending per FTE student",
  "FacSalary",  "Average monthly salary for full-time faculty",
  "FullTimeFac","Percent of faculty who are full-time",
  "Pell",       "Percent of students receiving Pell grants",
  "CompRate",   "Completion rate (percent who finish programme within 150% of normal time)",
  "Debt",       "Average debt for students who complete the programme",
  "Female",     "Percent of female students",
  "FirstGen",   "Percent of first-generation students",
  "MedIncome",  "Median family income (in $1,000)"
)




library(gt)
library(tibble)

college_dict |>
  gt() |>
  cols_label(
    Variable    = "Variable Name",
    Description = "Description"
  ) |>
  # stop wrapping in the header
  tab_style(
    style = cell_text(whitespace = "nowrap"),
    locations = cells_column_labels(columns = Variable)
  ) |>
  # bold headings
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) |>
  # bold variable names
  #tab_style(
  #  style = cell_text(weight = "bold"),
   # locations = cells_body(columns = Variable)
  #) |>
  # 👉 add horizontal space by widening the first column
  cols_width(
    Variable    ~ px(180),   # wider first column = more gap to the next
    Description ~ px(600)    # optional: set width for the description column
  ) |>
  opt_row_striping() |>
  tab_options(
    table.font.size  = px(13.6),
    data_row.padding = px(6)
  )
```




## 2 Worked Example

Consider the dataset provided in `datasets/HollywoodMovies.csv`,
containing **1295 observations** on the following **15 variables**:

```{r hollywood-dict, echo=FALSE, message=FALSE, warning=FALSE}

library(tibble)
library(gt)
library(dplyr)

hollywood_dict <- tribble(
  ~Variable,         ~Description,
  "Movie",           "Title of the movie",
  "LeadStudio",      "Primary U.S. distributor",
  "RottenTomatoes",  "Critics' rating (Rotten Tomatoes)",
  "AudienceScore",   "Audience rating (Rotten Tomatoes)",
  "Genre",           "Film genre (e.g., Action Adventure, Comedy, Thriller)",
  "TheatersOpenWeek","Number of screens on opening weekend",
  "OpeningWeekend",  "Opening weekend gross (in millions)",
  "BOAvgOpenWeekend","Average box office income per theatre, opening weekend",
  "Budget",          "Production budget (in millions)",
  "DomesticGross",   "U.S. gross income (in millions)",
  "ForeignGross",    "Foreign gross income (in millions)",
  "WorldGross",      "Worldwide gross income (in millions)",
  "Profitability",   "Worldwide gross as a percentage of budget",
  "OpenProfit",      "Percentage of budget recovered on opening weekend",
  "Year",            "Year of release"
)

hollywood_dict |>
  gt() |>
  cols_label(
    Variable    = "Variable Name",
    Description = "Description"
  ) |>
  # prevent header wrapping
  tab_style(
    style = cell_text(whitespace = "nowrap"),
    locations = cells_column_labels()
  ) |>
  # bold headers
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  # widen columns for spacing
  cols_width(
    Variable    ~ px(180),
    Description ~ px(600)
  ) |>
  opt_row_striping() |>
  tab_options(
    table.font.size  = px(13.6),
    data_row.padding = px(6)
  )
```

These data were compiled from Box Office Mojo, The Numbers, and Rotten
Tomatoes.

::: margin
We load the `tidyverse` package as we will use the functions\
`read_csv()` and `glimpse()` from this package.
:::

#### Load tidyverse and import the data

```{r load_library, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
```

::: margin
`read_csv()` reads CSV (comma-separated values) files.\
The loaded data are stored in an object called `movies` using the arrow
`<-`.
:::

```{r load_data, echo=TRUE, message=FALSE, warning=FALSE}
movies <- read_csv("HollywoodMovies.csv")
```

```{r}
movies <- movies |>
  select(1:15)

glimpse(movies)

```

### 3.1 Visualising categorical variables

| Variable       | Why categorical?                                 |
|----------------|--------------------------------------------------|
| **Movie**      | Unique labels (ID variable)                      |
| **LeadStudio** | Qualitative categories                           |
| **Genre**      | Qualitative categories                           |
| **Year**       | Distinct categories, not continuous numeric data |

You will create **bar plots** for:

-   `Genre`

-   `LeadStudio`

-   `Year`

Bar plots are the standard way to display categorical frequencies.

#### Genre frequency plot

```{r}
plot_genre <- ggplot(movies, aes(x = Genre)) +
  geom_bar() +
  labs(x = "Movie Genre", y = "Frequency") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
plot_genre
```

#### LeadStudio frequency plot

```{r}
library(tidyverse)
library(forcats)

plot_LeadStudio  <- ggplot(movies, aes(x = fct_infreq(LeadStudio))) +
  geom_bar() +
  labs(
    x = "Lead Studio (sorted by frequency)",
    y = "Count",
    title = "Frequency of Lead Studios"
  ) +
  scale_y_continuous(
    breaks = seq(0, 150, by = 20)   # More detailed count axis
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      size = 6,                    # Smaller labels so they fit
      lineheight = 0.8
    ),
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.y = element_text(size = 9)
  )

plot_LeadStudio
```

Flip

```{r, fig.height=15}
ggplot(movies, aes(x = fct_infreq(LeadStudio))) +
  geom_bar() +
  coord_flip() +
  labs(x = "Lead Studio (sorted by frequency)", y = "Count") +
  theme_minimal() +
  scale_y_continuous(breaks = seq(0, 150, by = 10))
```

#### Show only the top N studios (recommended)

Most analyses only need the *most frequent* studios.

```{r}
library(tidyverse)

top_studios <- movies |>
  count(LeadStudio) |>
  arrange(desc(n)) |>
  slice_head(n = 10)   # top 10 studios
top_studios
```

```{r}

# Plot with readable labels
ggplot(top_studios, aes(x = fct_reorder(LeadStudio, n), y = n)) +
  geom_col(fill = "grey40") +
  labs(
    x = "Lead Studio",
    y = "Count",
    title = "Top 10 Most Common Lead Studios"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1
    )
  )

```

-   **`slice(1:10)`** limits the graph to the top 10 studios → readable.

-   **`fct_reorder()`** orders bars by frequency.

-   **`angle = 45`** rotates labels so they don’t overlap.

-   **`base_size = 14`** makes everything easier to read.

If you want *horizontal bars* (even cleaner for long labels), use:

```{r}
ggplot(top_studios, aes(x = n, y = fct_reorder(LeadStudio, n))) +
  geom_col(fill = "grey40") +
  labs(
    x = "Count",
    y = "Lead Studio",
    title = "Top 10 Most Common Lead Studios"
  ) +
  theme_minimal(base_size = 14)

```

This horizontal version is often the best for categorical names.

#### Thinking question: Should we plot Movie Titles?

```{r}
ggplot(movies, aes(x = Movie)) + geom_bar()
```

::: {.callout-warning title="Would plotting movie titles give any useful insight?"}
\-`Movie` is an **identifier variable**: each title appears only once.
-A bar plot of titles would have **hundreds of bars of height 1**, which
is very hard to read and does not summarise the data. -In reports, we
usually **do not visualise ID variables**; instead, we focus on
variables that group observations into meaningful categories (such as
`Genre` or `LeadStudio`).
:::

#### Year frequency plot

```{r}
ggplot(movies, aes(x = factor(Year))) +
geom_bar() +
labs(
x = "Year of Release",
y = "Frequency",
title = "Number of Movies Released per Year"
) +
theme_minimal() +
theme(
axis.text.x = element_text(
angle = 45,
hjust = 1
)
)

```

### 3.2 Frequency tables with percentages

**Genre frequency table**

```{r}
library(tidyverse)
library(kableExtra)

# Genre frequency table
tbl_genre <- movies |>
  count(Genre, name = "n") |>
  mutate(
    Percent = round((n / sum(n)) * 100, digits = 2)
  ) |>
  arrange(desc(n))

tbl_genre

```

**LeadStudio frequency table**

```{r}
# LeadStudio frequency table
tbl_studio <- movies |>
  count(LeadStudio, name = "n") |>
  mutate(
    Percent = round((n / sum(n)) * 100, digits = 2)
  ) |>
  arrange(desc(n))

tbl_studio


```

```{r}
tbl_studio_top10 <- tbl_studio |>
  slice(1:10)
```

#### Formatting tables nicely

```{r}

kbl(
  list(
    tbl_genre,
    tbl_studio_top10
  ),
  booktabs = TRUE,
  digits = 2,
  col.names = c("Category", "Count", "Percent")
)

```

If you prefer **separate tables with captions** you can reference in the
text:

```{r}
# Genre table
tbl_genre |>
  kbl(
    booktabs = TRUE,
    digits = 2,
    caption = "Frequency distribution of movie genres",
    col.names = c("Genre", "Count", "Percent")
  )

# Lead studio table (top 10)
tbl_studio_top10 |>
  kbl(
    booktabs = TRUE,
    digits = 2,
    caption = "Top 10 most common lead studios",
    col.names = c("Lead Studio", "Count", "Percent")
  )

```

Then in your `.qmd` text you can reference them like:

``` markdown
Table \@ref(tab:genre-table) shows the frequency and percentage of each film genre.

Table \@ref(tab:studio-table) summarises the ten most common lead studios.
```

(using the corresponding chunk labels,
e.g. `{r genre-table}` and `{r studio-table}` for the chunks that create
each `kbl`).

### What is the most common category?

-   **Most common genre** = category with the highest frequency

-   **Most common lead studio** = category with the highest frequency

You can identify this from the **sorted** tables or by
using `slice_max()`:

```{r}
slice_max(tbl_genre, n, n = 1)
```

```{r}
slice_max(tbl_studio, n, n = 1)
```
