---
title: "Research Design & Data – MSDA IFP"
subtitle: "Week 08 – Tutorial 01 – Worked Solutions"
format:
  html:
    layout: article
    toc: true
    toc-depth: 3
    code-annotations: hover
    code-line-numbers: false
    code-fold: false
    code-block-background: true
    code-block-bg: true
    highlight-style: "pygments"
    css: styles.css
---

# Formative Report – Task 1 Solutions

## Data (CollegeScores Teaching Dataset)

In the tutorial you were asked to work with the **CollegeScores_teaching.csv** dataset.\
Below is one way to read the data and inspect its structure.

### 1. Reading the data into R

```{r}
library(tidyverse)

college <- read_csv("CollegeScores_teaching.csv")
```

### 2. View data ((View() / head())

```{r}
View(college)

```

Or look just at the first few rows:

```{r}
head(college)

```

### 3.Number of rows (observations)

```{r}
nrow(college) # number of rows 
```

### 4. Number of columns

```{r}
ncol(college)  # number of columns (variables)

dim(college)   # rows and columns together

```

### 5. Variable type in R (glimpse() / str() )

```{r}
glimpse(college)
```

### 6. Pick four variables, classify type + level, give example value

To see the type of each variable:

```{r}
summary(college)
```

This shows:

-   **Categorical variables**:\
    `Name`, `State`, `Control`, `Region`, `Locale`, `Main`

-   **Numeric variables (continuous)**:\
    `Enrollment`, `PartTime`, `TuitionIn`, `TuitionOut`

-   **Numeric variables (percentage/proportion)**:\
    `White`, `Black`, `Hispanic`, `Asian`, `Other`

### 6.1. Describing selected variables

Below are examples of how to describe four variables, covering both types.

#### **Categorical Variables**

```{r}
summary(college$Control)
summary(college$Region)
summary(college$Locale)

```

**Note:** Some variables such as *State*, *Region*, *Locale*, and *Control* are imported as character data. When stored as characters, `summary()` shows only their class and mode. After converting them to factors, `summary()` displays meaningful frequency counts for each category, which is the format used later when summarising categorical data.

**What to do?**

> Convert the variable to factor

```{r}
college$State <- factor(college$State)
summary(college$State)

```

```{r}
college$Region <- factor(college$Region)
summary(college$Region)
```

```{r}
college$Locale <- factor(college$Locale)
summary(college$Locale)
```

#### (7.) **Numeric variables**

Below are four example variables with type, level of measurement, and example values.

```{r}
summary(college$Enrollment)
summary(college$TuitionIn)
summary(college$TuitionOut)

```

### 8. Identify missing values and how many

-   Instruction suggests `summary(DATA)` and mentions NAs.

-   Worked solution talks about “minimal missing data” in the paragraph, but does not show explicit code. That is fine; if you want to mirror the hint a bit more closely you *could* add this tiny chunk to the worked solution:

```{r}
summary(college)
colSums(is.na(college))
```

`summary()` flags variables with `NA`s, and `colSums(is.na(college))` gives the exact number of missing values in each column. In this teaching file there are relatively few missing values across variables.

### 9. Example dataset description (Task 1 write-up)

A dataset containing information on 400 higher-education institutions in the United States was analysed. The dataset includes variables describing each institution’s geographical location (state, region, and local setting), sector type (public, private, or for-profit), and whether it represents a main or branch campus. Several numeric variables provide information on undergraduate enrolment, proportions of part-time students, and costs such as in-state and out-of-state tuition. The dataset also reports the demographic composition of each institution’s student body, including the percentage of students identifying as White, Black, Hispanic, Asian, or Other.

Initial inspection shows substantial variation in enrolment and tuition levels, and a small amount of missing data (for example, 1 missing value in *Locale*, 24 missing values in *Enrollment*, *PartTime*, *White*, *Black*, *Hispanic*, *Asian*, and *Other*, and 36 missing in *TuitionIn* and *TuitionOut*). Overall, the dataset offers a clear structure suitable for exploring relationships between institutional characteristics, costs, and student demographics.
